---
title: "Data Tidy 608 Final"
author: "Tyler Baker"
date: "5/16/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


# load data

```{r}
pisa_raw <- read.csv("https://raw.githubusercontent.com/tylerbaker01/Data-608-Final-Project/main/PISA_scores.csv")
gdp_raw <- read.csv("https://raw.githubusercontent.com/tylerbaker01/Data-608-Final-Project/main/worldGDP.csv")
lifeexp_raw <- read.csv("https://raw.githubusercontent.com/tylerbaker01/Data-608-Final-Project/main/LifeExpectancy.csv")
homicide_raw <- read.csv("https://raw.githubusercontent.com/tylerbaker01/Data-608-Final-Project/main/HomicdeRate.csv")
```

Since the pisa scores are only given every three years we will limit the data on the others to only the years 2006, 2009, 2012, 2015, 2018.
```{r}
gdp_raw <- subset (gdp_raw, select= c("Country.Name", "X2006", "X2009", "X2012", "X2015", "X2018"))
homicide_raw <- subset(homicide_raw, select=c("Country.Name", "X2006", "X2009", "X2012", "X2015", "X2018"))
lifeexp_raw <- subset(lifeexp_raw, select=c("Country.Name", "X2006", "X2009", "X2012", "X2015", "X2018"))
```

```{r}
colnames(gdp_raw) <- c("country", "2006", "2009", "2012", "2015", "2018")
colnames(homicide_raw) <- c("country", "2006", "2009", "2012", "2015", "2018")
colnames(lifeexp_raw) <- c("country", "2006", "2009", "2012", "2015", "2018")
```

Handeling NAs
```{r}
colSums(is.na(gdp_raw))
colSums(is.na(homicide_raw))
colSums(is.na(lifeexp_raw))
```


```{r}
gdp <- na.omit(gdp_raw)
homicide <- na.omit(homicide_raw)
lifeexp <- na.omit(lifeexp_raw)
```

```{r}
colnames(pisa_raw)<- c("year", "country", "reading", "reading se", "math", "math se", "science", "science se")
pisa <- pisa_raw[c("year", "country", "reading", "math", "science")]
```

There are two NAs labeled as "-" so I will fill them in with the international average which is 487. Which I calculated by taking the mean of the international test scores in reading
```{r}
pisa$reading <- gsub("-","487", pisa$reading)

```

Now I need to add gdp, homicide rate, and life expectancy to pisa.

First I need to filter pisa by year.
Second, I need to add a column for gdp, homicide rate, and life expectancy for country appear in both data sets.

Then I will be able to row bind the new datasets into one dataset.

```{r}
pisa_2018 <- pisa %>%
  filter(year==2018)
pisa_2015 <- pisa %>%
  filter(year==2015)
pisa_2012 <- pisa%>%
  filter(year==2012)
pisa_2009 <- pisa%>%
  filter(year==2009)
pisa_2006 <- pisa%>%
  filter(year==2006)
```

```{r}
pisa_2006$gdp <- gdp$`2006`[match(pisa_2006$country,gdp$country)]
pisa_2006$`life expectancy` <- lifeexp$`2006`[match(pisa_2006$country,lifeexp$country)]
pisa_2006$`homicide rate` <- homicide$`2006`[match(pisa_2006$country,homicide$country)]
pisa_2006 <- na.omit(pisa_2006)
pisa_2006$gdp <- as.numeric(pisa_2006$gdp)
pisa_2006$`life expectancy`<- as.numeric(pisa_2006$`life expectancy`)
pisa_2006$`homicide rate` <- as.numeric(pisa_2006$`homicide rate`)
pisa_2006$reading <- as.numeric(pisa_2006$reading)
pisa_2006$math <- as.numeric(pisa_2006$math)
pisa_2006$science <- as.numeric(pisa_2006$science)
avg_2006 <- c(2006, "International Average", mean(pisa_2006$reading, na.rm=TRUE), mean(pisa_2006$math, na.rm = TRUE), mean(pisa_2006$science, na.rm = TRUE), mean(pisa_2006$gdp, na.rm = TRUE), mean(pisa_2006$`life expectancy`, na.rm = TRUE), mean(pisa_2006$`homicide rate`, na.rm = TRUE))
pisa_2006<- rbind(pisa_2006,avg_2006)

pisa_2009$gdp <- gdp$`2009`[match(pisa_2009$country,gdp$country)]
pisa_2009$`life expectancy` <- lifeexp$`2009`[match(pisa_2009$country,lifeexp$country)]
pisa_2009$`homicide rate` <- homicide$`2009`[match(pisa_2009$country,homicide$country)]
pisa_2009 <- na.omit(pisa_2009)
pisa_2009$gdp <- as.numeric(pisa_2009$gdp)
pisa_2009$`life expectancy`<- as.numeric(pisa_2009$`life expectancy`)
pisa_2009$`homicide rate` <- as.numeric(pisa_2009$`homicide rate`)
pisa_2009$reading <- as.numeric(pisa_2009$reading)
pisa_2009$math <- as.numeric(pisa_2009$math)
pisa_2009$science <- as.numeric(pisa_2009$science)
avg_2009 <- c(2009, "International Average", mean(pisa_2009$reading, na.rm = TRUE), mean(pisa_2009$math, na.rm = TRUE), mean(pisa_2009$science, na.rm = TRUE), mean(pisa_2009$gdp, na.rm = TRUE), mean(pisa_2009$`life expectancy`, na.rm = TRUE), mean(pisa_2009$`homicide rate`, na.rm = TRUE))
pisa_2009<- rbind(pisa_2009,avg_2009)

pisa_2012$gdp <- gdp$`2012`[match(pisa_2012$country,gdp$country)]
pisa_2012$`life expectancy` <- lifeexp$`2012`[match(pisa_2012$country,lifeexp$country)]
pisa_2012$`homicide rate` <- homicide$`2012`[match(pisa_2012$country,homicide$country)]
pisa_2012 <- na.omit(pisa_2012)
pisa_2012$gdp <- as.numeric(pisa_2012$gdp)
pisa_2012$`life expectancy`<- as.numeric(pisa_2012$`life expectancy`)
pisa_2012$`homicide rate` <- as.numeric(pisa_2012$`homicide rate`)
pisa_2012$reading <- as.numeric(pisa_2012$reading)
pisa_2012$math <- as.numeric(pisa_2012$math)
pisa_2012$science <- as.numeric(pisa_2012$science)
avg_2012 <- c(2012, "International Average", mean(pisa_2012$reading, na.rm = TRUE), mean(pisa_2012$math, na.rm = TRUE), mean(pisa_2012$science, na.rm = TRUE), mean(pisa_2012$gdp, na.rm = TRUE), mean(pisa_2012$`life expectancy`, na.rm = TRUE), mean(pisa_2012$`homicide rate`, na.rm = TRUE))
pisa_2012<- rbind(pisa_2012,avg_2012)

pisa_2015$gdp <- gdp$`2015`[match(pisa_2015$country,gdp$country)]
pisa_2015$`life expectancy` <- lifeexp$`2015`[match(pisa_2015$country,lifeexp$country)]
pisa_2015$`homicide rate` <- homicide$`2015`[match(pisa_2015$country,homicide$country)]
pisa_2015 <- na.omit(pisa_2015)
pisa_2015$gdp <- as.numeric(pisa_2015$gdp)
pisa_2015$`life expectancy`<- as.numeric(pisa_2015$`life expectancy`)
pisa_2015$`homicide rate` <- as.numeric(pisa_2015$`homicide rate`)
pisa_2015$reading<- as.numeric(pisa_2015$reading)
pisa_2015$math <- as.numeric(pisa_2015$math)
pisa_2015$science <- as.numeric(pisa_2015$science)
avg_2015 <- c(2015, "International Average", mean(pisa_2015$reading, na.rm = TRUE), mean(pisa_2015$math, na.rm = TRUE), mean(pisa_2015$science, na.rm = TRUE), mean(pisa_2015$gdp, na.rm = TRUE), mean(pisa_2015$`life expectancy`, na.rm = TRUE), mean(pisa_2015$`homicide rate`, na.rm = TRUE))
pisa_2015<- rbind(pisa_2015,avg_2015)

pisa_2018$gdp <- gdp$`2018`[match(pisa_2018$country,gdp$country)]
pisa_2018$`life expectancy` <- lifeexp$`2018`[match(pisa_2018$country,lifeexp$country)]
pisa_2018$`homicide rate` <- homicide$`2018`[match(pisa_2018$country,homicide$country)]
pisa_2018 <- na.omit(pisa_2018)
pisa_2018$gdp <- as.numeric(pisa_2018$gdp)
pisa_2018$`life expectancy`<- as.numeric(pisa_2018$`life expectancy`)
pisa_2018$`homicide rate` <- as.numeric(pisa_2018$`homicide rate`)
pisa_2018$reading <- as.numeric(pisa_2018$reading)
pisa_2018$math <- as.numeric(pisa_2018$math)
pisa_2018$science<- as.numeric(pisa_2018$science)
avg_2018 <- c(2018, "International Average", mean(pisa_2018$reading, na.rm = TRUE), mean(pisa_2018$math, na.rm = TRUE), mean(pisa_2018$science, na.rm = TRUE), mean(pisa_2018$gdp, na.rm = TRUE), mean(pisa_2018$`life expectancy`, na.rm = TRUE), mean(pisa_2018$`homicide rate`, na.rm = TRUE))
pisa_2018<- rbind(pisa_2018,avg_2018)
```

```{r}
pisa <- rbind(pisa_2006,pisa_2009,pisa_2012,pisa_2015,pisa_2018)
```

```{r}
pisa<- na.omit(pisa)

pisa<- pisa[order(pisa$country, pisa$year),]

pisa<- pisa[c(2,1,3,4,5,6,7,8)]
```

```{r}
pisa <- pisa %>%
  pivot_longer(c("reading", "math", "science"), names_to = "subject", values_to = "score")
```

```{r}
write.csv(pisa, "C:\\Users\\tyler\\Downloads\\DATA 608\\pisa.csv")
```
